A Solution to Static vs. Dynamic Linking

### 2024-01-23 mss.tex
Post with updates.

\begin{abstract}
Computing systems that utilize library code must either link to a self-provided version (static linking) or a system-supplied version (dynamic linking).  This can lead to memory duplication or dependency hell.  The Urbit \texttt{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache.
\end{abstract}

@@

Nix

In Urbit, compilation means converting Hoon source code (as text) into Nock code (as a binary tree).  This process is handled by \texttt{++rash}, mint:ut

Linking, in the Urbit sense, derives from supplying nouns to nouns at compile time.\footnote{This is perhaps the biggest distinction from the conventional scenario for linking, which would refer to RAM words.  \emph{Ceteris paribus}, this article's discussion can inform the traditional dialogue.}  In the Urbit build arm \texttt{++ford}, a pair of builds becomes the build of a pair.  The subject (environment) used to build a file is the tuple \texttt{\string[import\_n … import\_2 import\_1 stdlib\string]}.  Since Hoon symbol lookup is left-to-right, this nests scopes seamlessly and predictably.  The linking technique is essentially trivial.

Ford compiles a Hoon source file into a data structure called a “vase”, a pair \texttt{\string[type noun\string]} where \texttt{noun} is a member of the set of nouns described by \texttt{type}.  Linking is thence just calling the \texttt{++slop} funciton, a one-liner from a pair of vases to a vase of the pair.

For example, the (trivial) Hoon source file with text \texttt{'3'} compiles to the vase \texttt{\string[\string[\%atom \%ud \textasciitilde\string] 3\string]}.  \texttt{3} is the value.  Its type is an atom (number), tagged as an unsigned decimal (\texttt{\%ud}) for printing.

As another example, the Hoon source file with text \texttt{'[3 0x4]'} compiles to \texttt{\string[\string[\%cell \string[\%atom \%ud \textasciitilde\string] \string[\%atom \%ux \textasciitilde\string]\string] \string[3 0x4\string]\string]}.  It's a vase of a \texttt{\%cell} (pair) of unsigned decimal number (\texttt{\%ud}) and unsigned hexadecimal number (\texttt{\%ux}).  If \texttt{/foo/hoon} is \texttt{'3'}, \texttt{/bar/hoon} is \texttt{'0x4'}, then importing both of those files changes the build subject to (a vase of) \texttt{\string[foo=3 bar=0x4 <stdlib>\string]}.

@@

In Urbit, everything is a “noun” (a binary tree with arbitrarily-sized integers at leaves).  If you “copy” a noun \texttt{foo}, like \texttt{\string[foo foo\string]}, the runtime just copies a pointer to it.  They are immutable, so everything shares structure.  Nouns are “persistent” data structures, like Clojure's collections.  If one copies a library, the copy is merely a pointer to the library—the library is a \emph{noun}.  If an imported library can be looked up from a build cache, the builder can copy it into a new app's build subject without duplicating it in memory.

@@

\section{The Modern \texttt{++ford} Build Cache}

@@

Since the caches weren't shared, they commonly held exactly the same data but generated independently.  For instance, the same library used on different desks would be built repeatedly and the memory was not shared unless the user manually ran the \texttt{|meld} command to manually deduplicate nouns.  For users with many apps installed, this added significant memory pressure.

@@

Reclaiming space from this cache becomes important.  Since this is a “true” cache, it's never incorrect to keep data in it.  One could adopt a heuristic such as to clear hourly or so you could use a heuristic such as "clear the whole thing every hour" or a least-recently-used (LRU) policy.  However, \texttt{++ford} has a long history of trying to use such heuristics and still using exorbitant amounts of memory.  The primary innovation of the former \texttt{++ford} cache was that its size is deterministic, and it stored no builds unless they could be generated from the head of its desk (\texttt{\%home} then \texttt{\%base}).

@@

\begin{enumerate}
  \item  Check if the build is in the per-desk cache; if so, return it
generate its dependencies.
  \item  Check if the build, with these dependencies, is in the global cache; if so, return it.
  \item  Else, build it.
\end{enumerate}
\noindent
In each case, the new build is added to either cache if it was not already prsent; and if it was not in the per-desk set of references, it is added there as well.

@@

Developers have long sought to balance the flexibility and portability of static linking with the better system demands of dynamic linking.  Despite care, the balancing act has led from well-managed mainframe code into the current linker spaghetti situation.  Urbit's \texttt{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache across desks.  This balances efficiency in code compilation and building with the reliability of solid-state computing.

>>>>

\begin{abstract}
Computing systems that utilize library code must either link to a self-provided version (static linking) or a system-supplied version (dynamic linking).  This can lead to memory duplication or dependency hell.  The Urbit \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache.
\end{abstract}

@@

Determinism has long been a desired characteristic of any given build system.  The source code should be a function of build environment and build instructions in a straightforward way.  Even accessing the linked libraries in a different order can alter the resulting binary, however, meaning that true reproducibility is elusive.  Declarative package managers like Nix \citep{Nix}~and Guix \citep{Guix}~use a functional package management approach to achieve reproducibility, marking packages using cryptographic hashes to track dependencies uniquely and repeatably.

In Urbit, compilation means converting Hoon source code (as text) into Nock code (as a binary tree).  This process is handled by \lstinline[style=inlinecode]{++rash}, \lstinline[style=inlinecode]{++mint:ut}, and other components of the Urbit build system.

Linking, in the Urbit sense, derives from supplying nouns to nouns at compile time.\footnote{This is perhaps the biggest distinction from the conventional scenario for linking, which refers to RAM words.  \emph{Ceteris paribus}, this article's discussion can inform the traditional dialogue.}  In the Urbit build arm \lstinline[style=inlinecode]{++ford}, a pair of builds becomes the build of a pair.  The subject (environment) used to build a file is the tuple \lstinline[style=inlinecode]{[import_n … import_2 import_1 stdlib]}.  Since Hoon symbol lookup is left-to-right, this nests scopes seamlessly and predictably.  The linking technique is essentially trivial.

Ford compiles a Hoon source file into a data structure called a “vase”, a pair \lstinline[style=inlinecode]{[type noun]} where \lstinline[style=inlinecode]{noun} is a member of the set of nouns described by \lstinline[style=inlinecode]{type}.  Linking is thence just calling the \lstinline[style=inlinecode]{++slop} funciton, a one-liner from a pair of vases to a vase of the pair.

For example, the (trivial) Hoon source file with text \lstinline[style=inlinecode]{'3'} compiles to the vase \lstinline[style=inlinecode]{[[%atom %ud ~] 3]}.  \lstinline[style=inlinecode]{3} is the value.  Its type is an atom (number), tagged as an unsigned decimal (\lstinline[style=inlinecode]{%ud}) for printing.

As another example, the Hoon source file with text \lstinline[style=inlinecode]{'[3 0x4]'} compiles to \lstinline[style=inlinecode]{[[%cell [%atom %ud ~] [%atom %ux ~]] [3 0x4]]}.  It's a vase of a \lstinline[style=inlinecode]{%cell} (pair) of unsigned decimal number (\lstinline[style=inlinecode]{%ud}) and unsigned hexadecimal number (\lstinline[style=inlinecode]{%ux}).  If \lstinline[style=inlinecode]{/foo/hoon} is \lstinline[style=inlinecode]{'3'}, \lstinline[style=inlinecode]{/bar/hoon} is \lstinline[style=inlinecode]{'0x4'}, then importing both of those files changes the build subject to (a vase of) \lstinline[style=inlinecode]{[foo=3 bar=0x4 <stdlib>]}.

@@

In Urbit, everything is a “noun” (a binary tree with arbitrarily-sized integers at leaves).  If you “copy” a noun \lstinline[style=inlinecode]{foo}, like \lstinline[style=inlinecode]{[foo foo]}, the runtime just copies a pointer to it.  They are immutable, so everything shares structure.  Nouns are “persistent” data structures, like Clojure's collections.  If one copies a library, the copy is merely a pointer to the library—the library is a \emph{noun}.  If an imported library can be looked up from a build cache, the builder can copy it into a new app's build subject without duplicating it in memory.

@@

\section{The Modern \lstinline[style=inlinecode]{++ford} Build Cache}

@@

Since the caches weren't shared, they commonly held exactly the same data but generated independently.  For instance, the same library used on different desks would be built repeatedly and the memory was not shared unless the user manually ran the \lstinline[style=inlinecode]{|meld} command to manually deduplicate nouns.  For users with many apps installed, this added significant memory pressure.

@@

Reclaiming space from this cache becomes important.  Since this is a “true” cache, it's never incorrect to keep data in it.  One could adopt a heuristic such as to clear hourly or so you could use a heuristic such as "clear the whole thing every hour" or a least-recently-used (LRU) policy.  However, \lstinline[style=inlinecode]{++ford} has a long history of trying to use such heuristics and still using exorbitant amounts of memory.  The primary innovation of the former \lstinline[style=inlinecode]{++ford} cache was that its size is deterministic, and it stored no builds unless they could be generated from the head of its desk (\lstinline[style=inlinecode]{%home} then \lstinline[style=inlinecode]{%base}).

@@

\begin{enumerate}
  \item  Check if the build is in the per-desk cache; if so, return it
generate its dependencies.
  \item  Check if the build, with these dependencies, is in the global cache; if so, return it.
  \item  Else, build it.
\end{enumerate}
\noindent
In each case, the new build is added to either cache if it was not already present; and if it was not in the per-desk set of references, it is added there as well.

@@

Developers have long sought to balance the flexibility and portability of static linking with the better system demands of dynamic linking.  Despite care, the balancing act has led from well-managed mainframe code into the current linker spaghetti situation.  Urbit's \lstinline[style=inlinecode]{++ford} build system elegantly solves for the linking problem by promoting structural sharing of objects (nouns) in memory and by utilizing a referentially transparent build cache across desks.  This balances efficiency in code compilation and building with the reliability of solid-state computing.
